---
title: "Analyzing Lebron James' Career Performance in the Regular Season and Playoffs"
output: html_document
date: "2026-04-02"
---
## Project Overview
This project analyzes LeBron James' offensive performance across his career, comparing regular season and playoff statistics. The goal is to identify whether his production changes in high-pressure playoff environments.

## Data & Tools
- Data sourced from compiled season statistics from basketball-reference.com
- Tools used: R, dplyr, tidyr, ggplot2

## Data Preparation
```{r, fig.width=10, fig.height=4, warning=FALSE}
# Load Libraries
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)

# Load Regular Season and Playoff Files
lebron <- read_csv("LebronYears.csv")
lebron <- type.convert(lebron, as.is = TRUE)

lebron_playoffs <- read_csv("lebron_career_playoffs.csv")
lebron_playoffs  <- type.convert(lebron_playoffs, as.is = TRUE)

# Convert selected columns to numeric and remove invalid rows
num_cols <- c("Age","G","GS","MP","FG","FGA","FG%","3P","3PA","3P%","2P","2PA","2P%","eFG%","FT","FTA","FT%","ORB","DRB","TRB","AST","STL","BLK","TOV","PF","PTS")
lebron[num_cols] <- lapply(lebron[num_cols], as.numeric)
lebron <- lebron[!is.na(lebron$Age) & lebron$Age != 41,]

lebron_playoffs[num_cols] <- lapply(lebron_playoffs[num_cols], as.numeric)
lebron_playoffs <- lebron_playoffs[1:20,]
lebron_playoffs$Season[14] <- "2018-19"
lebron_playoffs$Season[17] <- "2021-22"

# Give an identifier to each data frame
lebron$type <- "Regular Season"
lebron_playoffs $type <- "Playoffs"

# Build a combined data frame
lebron_combined <- bind_rows(lebron, lebron_playoffs)

championship_years <- c("2011-12", "2012-13", "2015-16", "2019-20")
lebron_combined <- lebron_combined |>
  mutate(Champion = ifelse(Season %in% championship_years & type == "Playoffs", "Champion", "Other"))

lebron_combined$PRA <- lebron_combined$PTS + lebron_combined$AST + lebron_combined$TRB

# Reshape scoring, rebounding, and assist data into long format for plotting
lebron_long <- lebron |>
  select(Season,PTS,AST,TRB)|>
  pivot_longer(
    cols = c(PTS,AST,TRB),
    names_to = "Stat",
    values_to = "Value"
  )
lebron_long <- lebron_long |>
  mutate(Stat = recode(Stat, "TRB" = "REB"))
lebron_long$Stat <- factor(lebron_long$Stat,
                           levels = c("PTS", "REB", "AST"))

lebron_long_po <- lebron_playoffs |>
  select(Season,PTS,AST,TRB)|>
  pivot_longer(
    cols = c(PTS,AST,TRB),
    names_to = "Stat",
    values_to = "Value"
  )
lebron_long_po <- lebron_long_po |>
  mutate(Stat = recode(Stat, "TRB" = "REB"))
lebron_long_po$Stat <- factor(lebron_long_po$Stat,
                           levels = c("PTS", "REB", "AST"))

# Build the team eras dataframe
team_eras <- data.frame(
  xmin = c(0.5, 7.5, 11.5, 16.5),
  xmax = c(7.5, 11.5, 16.5, 22.5),
  ymin = -Inf,
  ymax = Inf,
  Team = c("CLE1", "MIA", "CLE2", "LAL")
)
```
Cleaned and combined datasets to allow direct comparison between regular season and playoff performance. 

## Career Performance Overview

### Regular Season Trends
```{r, warning=FALSE}
# Create regular season PTS/REB/AST trend plot
lebrons_career <-ggplot(lebron_long, aes(x = Season, y = Value, color = Stat, group = Stat)) +
  scale_color_manual(values = c("PTS" = "#FDB927", "REB" = "#2F9E44", "AST" = "#9D4EDD"))+
  geom_line(linewidth = 1.25) +
  geom_point(size = 1.5)+
  labs(title = "LeBron James - Points, Rebounds, and Assists Per Game by Season",
       x = "Season", 
       y = "Points, Rebounds, Assists") +
    theme_minimal() +
    expand_limits(y = 0)+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(
    limits = c(0,35),
    breaks = seq(0, 30, by = 5))

lebrons_career
```

This chart shows how LeBron's offensive production evolved across his career during the regular season.

```{r, warning=FALSE}
# Create playoff PTS/REB/AST trend plot
lebrons_playoff_career <-ggplot(lebron_long_po, aes(x = Season, y = Value, color = Stat, group = Stat)) +
  scale_color_manual(values = c("PTS" = "#FDB927", "REB" = "#2F9E44", "AST" = "#9D4EDD"))+
  geom_line(linewidth = 1.25) +
  geom_point(size = 1.5)+
  labs(title = "LeBron James - Points, Rebounds, and Assists Per Playoff Game by Season",
       x = "Playoff Season", 
       y = "Points, Rebounds, Assists") +
    theme_minimal() +
    expand_limits(y = 0)+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(
    limits = c(0,36),
    breaks = seq(0, 30, by = 5))

lebrons_playoff_career
```

This chart shows LeBron's offensive playoff production from year to year.

## Metric Comparison

### Points Per Game (PPG)

```{r, warning=FALSE}
# Builds PPG Plot
lebron_ppg <-ggplot(lebron_combined, aes(x = Season, y = PTS, group = type)) +
  geom_line(aes(color = type), size = 2) +
  geom_point(aes(color = type), size = 1.33)+
  geom_point(
  data = lebron_combined %>% filter(Champion == "Champion"),
  color = "gold",
  size = 4,
  shape = 17
)+
  labs(title = "LeBron James - Points Per Game Regular Season vs Playoffs",
       x = "Season", 
       y = "Points Per Game") +
    theme_minimal() +
    expand_limits(y = 0)+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(
    limits = c(20,36),
    breaks = seq(0, 30, by = 10))

lebron_ppg
```

### Assists Per Game (APG)

```{r, warning=FALSE}
# Builds AST Plot
lebron_ast <-ggplot(lebron_combined, aes(x = Season, y = AST, group = type)) +
  geom_line(aes(color = type), size = 2) +
  geom_point(aes(color = type), size = 1.33)+
  geom_point(
  data = lebron_combined %>% filter(Champion == "Champion"),
  color = "gold",
  size = 4,
  shape = 17
)+
  #geom_text(aes(label = AST), vjust = -1, size = 2.5)+
  labs(title = "LeBron James - Assists Per Game Regular Season vs Playoffs",
       x = "Season", 
       y = "Assists Per Game") +
    theme_minimal() +
    expand_limits(y = 0)+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(
    limits = c(0,15),
    breaks = seq(0, 30, by = 5))

lebron_ast
```

### Rebounds Per Game (RPG)

```{r, warning=FALSE}
# Builds REB Plot
lebron_rebs <-ggplot(lebron_combined, aes(x = Season, y = TRB, group = type)) +
  geom_line(aes(color = type), size = 2) +
  geom_point(aes(color = type), size = 1.33)+
  geom_point(
  data = lebron_combined %>% filter(Champion == "Champion"),
  color = "gold",
  size = 4,
  shape = 17
)+
  #geom_text(aes(label = TRB), vjust = -1, size = 2.5)+
  labs(title = "LeBron James - Rebounds Per Game Regular Season vs Playoffs",
       x = "Season", 
       y = "Rebounds Per Game") +
    theme_minimal() +
    expand_limits(y = 0)+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(
    limits = c(0,15),
    breaks = seq(0, 30, by = 5))

lebron_rebs
```

## Points, Rebounds, Assists (PRA)

```{r, warning=FALSE}
# Builds PRA Plot
lebron_pra <-ggplot(lebron_combined, aes(x = Season, y = PRA, group = type)) +
  geom_line(aes(color = type), size = 2) +
  geom_point(aes(color = type), size = 1.33)+
  geom_point(
  data = lebron_combined %>% filter(Champion == "Champion"),
  color = "gold",
  size = 4,
  shape = 17
)+
  labs(title = "LeBron James - Points, Rebounds, and Assists Per Game Regular Season vs Playoffs",
       x = "Season", 
       y = "PRA Per Game") +
    theme_minimal() +
    expand_limits(y = 0)+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(
    limits = c(30,60),
    breaks = seq(0, 60, by = 10))

lebron_pra
```

## Performance Differences
To directly compare playoff and regular season performance, I created a differences data frame by subtracting regular season averages from playoff averages for each season. Positive values indicate better playoff performance, while negative values indicate lower production in the playoffs.

```{r, warning=FALSE}

# Building the LeBron Playoff and Regular Season Data Frame
lebron_pts_diff <- lebron_combined |>
  group_by(Season,type)|>
  summarize(PTS = mean(PTS, na.rm = TRUE), .groups = "drop")|>
  pivot_wider(names_from = type, values_from = PTS)

lebron_reb_diff <- lebron_combined |>
  group_by(Season,type)|>
  summarize(TRB = mean(TRB, na.rm = TRUE), .groups = "drop")|>
  pivot_wider(names_from = type, values_from = TRB)

lebron_ast_diff <- lebron_combined |>
  group_by(Season,type)|>
  summarize(AST = mean(AST, na.rm = TRUE), .groups = "drop")|>
  pivot_wider(names_from = type, values_from = AST)

# Remove NAs and calculate the difference between playoff and regular season
lebron_pts_diff <- lebron_pts_diff[!is.na(lebron_pts_diff$Playoffs),]
lebron_pts_diff$Playoffs[is.nan(lebron_pts_diff$Playoffs)] <- NA
lebron_pts_diff$pts_difference = lebron_pts_diff$Playoffs - lebron_pts_diff$`Regular Season`

lebron_reb_diff <- lebron_reb_diff[!is.na(lebron_reb_diff$Playoffs),]
lebron_reb_diff$Playoffs[is.nan(lebron_reb_diff$Playoffs)] <- NA
lebron_reb_diff$reb_difference = lebron_reb_diff$Playoffs - lebron_reb_diff$`Regular Season`

lebron_ast_diff <- lebron_ast_diff[!is.na(lebron_ast_diff$Playoffs),]
lebron_ast_diff$Playoffs[is.nan(lebron_ast_diff$Playoffs)] <- NA
lebron_ast_diff$ast_difference = lebron_ast_diff$Playoffs - lebron_ast_diff$`Regular Season`

# Rename the columns
lebron_pts_diff <- lebron_pts_diff |>
  rename(PTS_RS = `Regular Season`,
         PTS_PO = Playoffs)

lebron_reb_diff <- lebron_reb_diff |>
  rename(REB_RS = `Regular Season`,
         REB_PO = Playoffs)

lebron_ast_diff <- lebron_ast_diff |>
  rename(AST_RS = `Regular Season`,
         AST_PO = Playoffs)

lebron_diff <- lebron_pts_diff |>
  full_join(lebron_reb_diff, by = "Season")|>
  full_join(lebron_ast_diff, by = "Season")

# Adding PRA line

lebron_diff$pra_rs <- lebron_diff$PTS_RS + lebron_diff$REB_RS + lebron_diff$AST_RS
lebron_diff$pra_po <- lebron_diff$PTS_PO + lebron_diff$REB_PO + lebron_diff$AST_PO
lebron_diff$pra_difference <- lebron_diff$pra_po - lebron_diff$pra_rs

```

## Average Difference By Metric
```{r}
pts_diff <- mean(lebron_diff$pts_difference)
reb_diff <-mean(lebron_reb_diff$reb_difference)
ast_diff <-mean(lebron_diff$ast_difference)
pra_diff <- mean(lebron_diff$pra_difference)
```
On average, LeBron performed better in the playoffs, scoring 1.07 more points and grabbing 1.24 more rebounds. However, his assists decreased slightly by 0.33, while overall production (PRA) increased by 1.99.

## Statistical Testing
To test whether these observed differences were statistically significant, I used paired t-tests comparing LeBron's regular season and playoff performance across seasons.

### Points Paired T-test
```{r}
# Running Paired T-Tests
# Pts T-Test

# Histogram of Differences
hist(lebron_diff$pts_difference,
     main = "Differences in Regular Season and Playoff PPG",
     xlab = "Difference in PPG")

pts_ttest <- t.test(lebron_diff$pts_difference, mu = 0)
```
**Points Interpretation:**

On average, LeBron James scores approximately 1.07 more points per game in the playoffs compared to the regular season. However, a paired t-test indicates that this difference is not statistically significant (t = 1.34, p = 0.198), indicating the increase is not consistent across seasons.

### Rebounds Paired T-test
```{r}
hist(lebron_diff$reb_difference,
     main = "Differences in Regular Season and Playoff Rebounds",
     xlab = "Difference in REB")

reb_ttest <- t.test(lebron_diff$reb_difference, mu = 0)
```
**Rebounds Interpretation:**

In contrast, LeBron averages 1.24 more rebounds per game in the playoffs than in the regular season. This difference is statistically significant (t = 3.83, p = 0.001), indicating a consistent increase in rebounding performance during postseason play.

### Assists Paired T-test
```{r}
hist(lebron_diff$ast_difference,
     main = "Differences in Regular Season and Playoff Assists",
     xlab = "Difference in AST")

ast_ttest <- t.test(lebron_diff$ast_difference, mu = 0)
```
**Assists Interpretation:**

LeBron averages 0.33 fewer assists per game in the playoffs compared to the regular season. This difference is not statistically significant (t = -1.28, p = 0.217), suggesting no consistent change in playmaking output.

### PRA Paired T-test
```{r}
hist(lebron_diff$pra_difference,
     main = "Differences in Regular Season and Playoff Points, Rebounds, and Assists",
     xlab = "Difference in PRA")

pra_ttest <- t.test(lebron_diff$pra_difference, mu = 0)
```
**Points, Rebounds, Assists (PRA) Interpretation:**

Overall, LeBron averages 1.99 more combined points, rebounds, and assists (PRA) in the playoffs. However, this difference is not statistically significant (t = 2.03, p = 0.0587), barely missing the conventional significance threshold of α = 0.05.

## Key Findings

Overall, the analysis shows that LeBron’s performance remains largely consistent between the regular season and playoffs. While scoring, assists, and overall PRA do not show statistically significant changes, rebounding demonstrates a clear and consistent increase in the postseason. This suggests that while LeBron maintains a high level of performance year-round, he contributes additional impact in rebounding during playoff competition.

- **Scoring (PPG):** Increased slightly in the playoffs, but not statistically significant.
- **Rebounding (RPG):** Increased in the playoffs and was statistically significant, indicating a consistent boost in rebounding impact.
- **Assists (APG):** Decreased slightly in the playoffs, but not statistically significant.
- **Overall Offensive Production (PRA):** Increased modestly, but did not reach statistical significance.

## Limitations
While this analysis provides useful insights, several limitations should be considered:

- Dataset required manual cleaning and adjustments, which may introduce mionr inconsistencies.
- The analysis uses a relatively small sample size (~20 Seasons), limiting statistical power.
- Only used basic box score metrics (points, rebounds, assists)
- Contextual factors such as team composition, opponent strength, and role changes were not considered

## Future Work
There are several opportunities to expand and improve this analysis:

- Incorporate advanced metrics such as true shooting percentage, usage rate, effective field goal percentage, or player efficiency rating.
- Compare LeBron's playoff performance to other all-time great players to provide broader context.
- Analyze individual playoff runs at the series level.
- Build an interactive dashboard to allow users to explore performance trends dynamically. 
